---
title: "Your first prediction report"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your first prediction report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

AutoXplainR fits a tabular prediction model, evaluates it against a simple
baseline, and retains fitted explanations and an optional HTML report. The
ordinary workflow runs locally. This example predicts a parcel's delivery time
using information available at dispatch. All data below are synthetic.

## Choose what can be known when predicting

```{r data}
set.seed(2026)
parcels <- data.frame(
  parcel_id = sprintf("parcel-%03d", 1:240),
  dispatched = as.Date("2025-01-01") + 0:239,
  distance_km = runif(240, 20, 600),
  service = rep(c("standard", "express"), 120)
)
parcels$delivery_hours <- 12 + parcels$distance_km / 25 -
  5 * (parcels$service == "express") + rnorm(240, sd = 2)
parcels$delivered <- parcels$dispatched + parcels$delivery_hours / 24
parcels$distance_km[c(8, 40, 90, 211)] <- NA_real_

# Exclude the identifier and the delivery timestamp, which is only known later.
inputs <- parcels[c("dispatched", "distance_km", "service", "delivery_hours")]
```

The package cannot decide whether a column leaks future information. Choose
predictors from the intended prediction task before looking at model scores.
Identifiers also require a decision: an arbitrary parcel ID should not predict
delivery time. Repeated customers or sites would need a grouped validation plan.

## Reserve later observations and inspect preprocessing

```{r fit}
result <- autoxplain(
  inputs, "delivery_hours",
  validation = validation_split(time = "dispatched"),
  preprocessing_config = list(novel_level_strategy = "error"),
  seed = 2026
)
result

recipe <- result$preprocessing_metadata$training_data$recipe
recipe$final_columns
recipe$imputations
recipe$factor_levels
recipe$novel_level_strategy
```

This fits a linear regression and an intercept-only baseline on the earlier
rows. The latest 20% of distinct dispatch dates are reserved for evaluation.
`dispatched` determines the split and is excluded from predictors. Missing
predictor values are imputed using values learned on training rows; character
predictors are converted to factors using their training levels.

The default novel-category strategy is `"mode"`, which maps an unseen category
to the most frequent training category. Here we deliberately choose `"error"`
so that a new service cannot be silently treated as an existing one.

## Read the result before explaining it

```{r evaluate}
result$leaderboard
result$evaluation$metric_definitions
result$evaluation$notes
```

RMSE summarizes prediction error in hours, giving larger errors more weight.
The baseline predicts the training mean without using distance or service.
Beating it is useful context, but does not establish that the error is acceptable
for a delivery promise. The scores describe these reserved rows; transfer to a
future population needs further evidence.

```{r explanations}
result$explanations$audit$importance
result$explanations$audit$findings
result$explanations$failures
```

Permutation importance describes how the fitted model's evaluation loss changes
when an input is shuffled. Its intervals describe shuffle Monte Carlo error,
not uncertainty about a population. ALE and PDP summarize fitted associations,
not the effect of changing a real parcel's service. A check that was not run or
could not be computed remains visible with its reason.

## Predict new parcels and share the report

```{r predict}
new_parcels <- data.frame(distance_km = c(100, NA), service = c("express", "standard"))
predict(result, new_parcels)

unknown_service <- data.frame(distance_km = 100, service = "cargo")
tryCatch(predict(result, unknown_service), error = function(error) conditionMessage(error))
```

Prediction applies the saved training recipe and preserves input row order.
An outcome column is unnecessary. The unknown service produces an explicit
error under our chosen strategy; decide how that category should be represented
before refitting or changing the recipe.

```{r report}
path <- tempfile(fileext = ".html")
memo <- generate_natural_language_report(result)
render_model_report(result, path, narrative = memo)
file.exists(path)
unlink(path)
```

Use a persistent path such as `"delivery-report.html"` to keep the report. It
opens offline in a browser. Review feature names and diagnostics before sharing.
`saveRDS(result, "analysis.rds")` retains models **and training/evaluation data**.

For classification, set outcome factor levels deliberately: binary probabilities
refer to the **second** level. For example, `factor(outcome, levels = c("no", "yes"))`
makes predictions probabilities of `"yes"`. Numeric outcomes with exactly two
values default to classification; set `task = "regression"` for a numeric target.

Continue with [model selection](model-selection.html),
[validation and diagnostics](validation-and-diagnostics.html),
[existing models and narratives](existing-models-and-narratives.html), or the
[statistical methods](statistical-methods.html).
