---
title: "Auditing Explanation Evidence with AutoXplainR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auditing Explanation Evidence with AutoXplainR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

## From explanation output to evidence

Permutation importance and feature-effect plots are useful descriptions of a
fitted model. They are also easy to over-interpret. A feature can appear
important because of a particular random permutation, correlated predictors can
make marginal perturbations unrealistic, and a different but equally accurate
model can support a different story.

AutoXplainR treats an explanation as a claim that needs a recorded evidence
contract:

1. Which fitted model and evaluation distribution are being described?
2. Does the numerical result survive repetition of its random computation?
3. Are feature-dependence assumptions under pressure?
4. Do similarly performing supplied models support the same ranking?
5. What language is justified by those diagnostics?

The resulting grade is a transparent triage heuristic. It is not a p-value,
causal result, fairness assessment, or certification.

## Use held-out data

This small example uses a fixed split only to keep the vignette reproducible.
A real analysis should design its evaluation split before fitting and retain
enough observations for stable diagnostics.

```{r fit}
train <- mtcars[1:24, ]
test <- mtcars[25:32, ]

model_a <- lm(mpg ~ wt + hp + disp, data = train)
model_b <- lm(mpg ~ wt + hp + qsec, data = train)
```

`explain_model()` validates how each model produces predictions. The explainer
stores feature data, outcomes, task metadata, and a lightweight provenance ID.

```{r explainers}
explainer_a <- explain_model(
  model_a,
  test,
  y = "mpg",
  label = "displacement model",
  metadata = list(evaluation_role = "test")
)

explainer_b <- explain_model(
  model_b,
  test,
  y = "mpg",
  label = "acceleration model",
  metadata = list(evaluation_role = "test")
)

explainer_a
```

For an unsupported model class, pass `predict_function`. Regression functions
return numeric predictions; binary functions return the positive-class
probability; multiclass functions return a named probability matrix.

## Inspect repeated permutation importance

```{r importance}
importance <- calculate_permutation_importance(
  explainer_a,
  features = c("wt", "hp", "disp"),
  n_repeats = 10,
  seed = 2026
)

importance
```

The result deliberately retains negative values. Its interval is a Monte Carlo
interval across permutation repeats. Repeating an algorithm does not create new
independent observations from a population, so this is not population-level
inference.

Feature groups can be permuted jointly:

```{r groups}
calculate_permutation_importance(
  explainer_a,
  feature_groups = list(engine = c("hp", "disp"), weight = "wt"),
  n_repeats = 5,
  seed = 2026
)
```

The optional `within` argument permutes within explicit strata. This is useful
as a blocked sensitivity analysis, but it is not advertised as a general
conditional permutation estimator.

## Prefer ALE for dependent predictors

Partial dependence replaces one feature while leaving all others unchanged.
With dependent predictors, that can ask the model to score combinations absent
from the reference data. Accumulated local effects (ALE) instead average local
prediction differences within observed feature bins (Apley and Zhu, 2020).

```{r effects}
ale <- explain_effect(explainer_a, feature = "wt")
pdp <- explain_effect(explainer_a, feature = "wt", method = "pdp")

ale
attr(pdp, "max_association")
attr(pdp, "dependence_warning")
```

Both effect outputs include relative empirical support. Their uncertainty
columns remain descriptive computation diagnostics.

## Audit several plausible models

```{r audit}
audit <- audit_explanations(
  list(explainer_a, explainer_b),
  features = c("wt", "hp", "disp", "qsec"),
  n_repeats = 10,
  performance_tolerance = 0.10,
  dependence_threshold = 0.70,
  seed = 2026
)

audit
```

The near-optimal indicator is relative to the models supplied to this call. It
does not claim to enumerate every near-optimal model in a hypothesis class. If
near-optimal supplied models disagree, the audit recommends reporting an
importance range rather than a single-model ranking.

The main components are ordinary data structures:

```{r components}
audit$performance
audit$dependence
audit$explanation_agreement$importance_ranges
audit$findings
```

Evidence grades control the wording that the report permits:

- **A:** stable marginal evidence under the configured diagnostics;
- **B:** usable with the stated uncertainty;
- **C:** sensitivity finding only;
- **D:** do not make a feature claim.

These cutoffs are visible in the source and should be configured and justified
for the application. A grade cannot turn an associational explanation into a
causal one.

## Preserve the review artifact

```{r report}
path <- tempfile(fileext = ".html")
render_explanation_report(audit, path)
file.exists(path)
unlink(path)
```

The report is a standalone HTML file with findings, action items, model and
feature tables, interpretation boundaries, configuration, timestamps, and
explainer IDs. Store it with the model version, evaluation-data version, and
analysis code.

## Optional H2O AutoML adapter

The explanation core does not require H2O or Java. If H2O is useful for fitting
a candidate set, keep it at the boundary:

```{r h2o, eval=FALSE}
result <- autoxplain(
  training_data,
  target_column = "outcome",
  test_data = held_out_data,
  max_models = 10,
  max_runtime_secs = 300,
  seed = 2026
)

explainers <- as_explainers(result)
audit <- audit_explanations(explainers)
generate_dashboard(result, "automl-evidence.html")
```

Held-out test data are not passed to H2O as a validation frame by default.
Numeric two-level outcomes are classified correctly, and training
preprocessing is applied to evaluation data through a fitted recipe.

## Optional narratives

```{r narrative}
memo <- generate_natural_language_report(audit, use_remote = FALSE)
cat(substr(memo, 1, 300), "...")
```

The local narrative is deterministic. Remote Gemini generation is an explicit
optional boundary and receives only aggregated diagnostics. The prompt forbids
causal, safety, fairness, and compliance claims, but generated text still
requires human review.

## Literature context

Permutation-based model reliance and model-class reliance are discussed by
Fisher, Rudin, and Dominici (2019). ALE and the extrapolation problem of PDPs
under dependent features are developed by Apley and Zhu (2020). AutoXplainR
operationalizes these concerns as a review workflow; it does not claim a new
formal inferential estimator.
