---
title: "Your First Model with AutoXplainR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your First Model with AutoXplainR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

## What AutoXplainR is for

Fitting a model is only one part of a useful analysis. A first-time modeler also
needs to know whether the model works on unseen rows, whether it improves on a
simple guess, which fitted patterns matter, and how to describe those patterns
without claiming more than the analysis established.

AutoXplainR puts those questions in one workflow. Its safe default is local and
requires neither Java nor an API account. This vignette uses the small built-in
`mtcars` data only so every step is reproducible.

## Fit and evaluate in one call

```{r guided-fit}
result <- autoxplain(
  mtcars,
  target_column = "mpg",
  seed = 2026
)

result
```

`mpg` is numeric with more than two distinct values, so this is recognized as a
regression problem. AutoXplainR reproducibly holds out 20% of the rows, learns
preprocessing from the remaining rows, and fits two models:

- a linear regression using the input features; and
- an intercept-only baseline that ignores all input features.

The baseline matters. It answers whether the fitted relationships improve on a
very simple reference rather than merely producing predictions.

## Read the held-out metrics

```{r evaluation}
result$leaderboard
result$evaluation$metric_definitions
result$evaluation$improvement_over_baseline
```

For regression, RMSE is the primary comparison metric. It is a typical error in
the outcome's units, with larger errors weighted more heavily. Lower is better.
MAE is the average absolute error. R-squared compares squared error with a
held-out mean reference and can be negative on genuinely unseen data.

The exact values from a six-row `mtcars` holdout are not a basis for scientific
conclusions; the example demonstrates the contract. Real analyses need an
evaluation set large and representative enough for the intended use.

## Build the report

```{r report}
path <- tempfile(fileext = ".html")
render_model_report(result, path, top_features = 4, n_repeats = 5)
file.exists(path)
unlink(path)
```

The standalone report uses progressive disclosure:

1. the modeling question and baseline verdict;
2. every held-out metric with definitions;
3. input reliance and fitted effect directions;
4. warnings and concrete next actions; and
5. a collapsed technical evidence audit and complete provenance.

The report does not need an LLM, Plotly, H2O, or a browser runtime after it is
written.

## Compare a small candidate set

The first call stays intentionally simple. When comparing a few approachable
model shapes would help, use the explicit comparison mode:

```{r comparison}
comparison <- autoxplain(
  iris,
  target_column = "Sepal.Length",
  model_set = "comparison",
  test_fraction = 0.4,
  seed = 2026
)

model_tradeoffs(comparison)
```

This adds a shallow tree and a more flexible tree to the pre-specified
statistical model and intercept-only baseline. Pareto status asks whether
another supplied model is at least as good on both held-out performance and
model size, with a strict improvement on one.

The comparison is descriptive. AutoXplainR does not silently promote the
holdout winner to primary, because selecting on the holdout and reporting the
same score as final performance would be optimistic. Use fresh validation or a
pre-specified resampling procedure when model selection itself is the goal.

An interactive view is available when Plotly is installed:

```{r comparison-plot, eval=FALSE}
plot_model_comparison(comparison)
```

## Investigate one fitted pattern

The guided result can be converted to a model-agnostic explainer. Select the
primary model rather than the simple baseline:

```{r explainer}
explainer <- as_explainers(result, models = "main_model")$main_model
explainer
```

Repeated permutation importance measures how much held-out performance changes
when an input is shuffled:

```{r importance}
importance <- calculate_permutation_importance(
  explainer,
  features = c("wt", "hp", "disp"),
  n_repeats = 10,
  seed = 2026
)

importance
```

The result keeps every repeat, a Monte Carlo interval, and sign stability. The
interval describes randomness from shuffling. It is not a confidence interval
for a population and cannot support a causal claim.

`explain_effect()` defaults to accumulated local effects (ALE) for numeric
inputs:

```{r effects}
weight_effect <- explain_effect(explainer, feature = "wt")
weight_effect
```

ALE summarizes local prediction changes inside observed feature bins. It
reduces the most direct extrapolation problem of a marginal partial-dependence
plot under dependent inputs, but it still describes a fitted prediction
function—not the effect of intervening on vehicle weight.

## Open the reliability layer

The evidence audit is available when more technical review is useful:

```{r audit}
audit <- audit_explanations(
  as_explainers(result),
  features = c("wt", "hp", "disp"),
  n_repeats = 10,
  seed = 2026
)

audit
```

The audit checks computation stability, feature association, supplied-model
performance, and explanation agreement. Because this example supplies a
primary model and a baseline, it can also demonstrate that only competitively
performing supplied models should influence an explanation comparison.

The A--D grade is a triage heuristic. It is not a p-value, certification,
fairness test, or safety assessment.

## Add a narrative only if it helps communication

The deterministic provider works offline:

```{r narrative}
memo <- generate_natural_language_report(result)
cat(substr(memo, 1, 400), "...")
attr(memo, "narrative_provenance")
```

Remote use requires an explicit provider:

```{r remote-narrative, eval=FALSE}
Sys.setenv(GEMINI_API_KEY = "...")
memo <- generate_natural_language_report(result, provider = "gemini")

# Or use a model running locally through Ollama
memo <- generate_natural_language_report(result, provider = "ollama")
```

Only aggregated diagnostics are placed in the prompt. Raw rows, fitted objects,
case-level predictions, and secrets are excluded. Provider and model details
are attached to the returned text so the prose can be audited. Generated prose
still needs human review against the numerical report.

## Classification uses the same contract

```{r multiclass}
flowers <- autoxplain(iris, target_column = "Species", seed = 2026)
flowers
flowers$evaluation$metric_definitions
```

Classification splits are stratified. Binary tasks report log loss, Brier
score, a binned calibration gap, accuracy, balanced accuracy, and ROC AUC.
Multiclass tasks report log loss, Brier score, predicted-class calibration,
accuracy, and macro recall. Log loss is primary because it evaluates
probability quality and penalizes confident wrong predictions.

Calibration asks a more literal question: when the model reports a probability
near 70%, does the corresponding event happen about 70% of the time on the
held-out rows? The report explains this visually, and the grouped values remain
available directly:

```{r calibration}
calibration_diagnostics(flowers)
```

The binned gap is descriptive. It changes with the evaluation sample and the
grouping, so it should be read beside log loss and Brier score rather than as a
population guarantee.

## Bring an existing model

The explanation layer does not depend on the guided fitting engine. Wrap an
ordinary R model directly:

```{r existing-model}
train <- mtcars[1:24, ]
test <- mtcars[25:32, ]
fit <- lm(mpg ~ wt + hp + disp, data = train)

existing <- explain_model(
  fit,
  test,
  y = "mpg",
  label = "existing linear model",
  metadata = list(evaluation_role = "test")
)
```

Other frameworks can supply a `predict_function`. H2O AutoML remains an
explicit optional fitting engine through `autoxplain(..., engine = "h2o")`.

## What to say—and not say

Reasonable language is specific to the fitted model and evaluation data:

> On the held-out rows, shuffling `wt` changed the fitted model's RMSE by the
> reported amount, with the displayed variation across permutation repeats.

The following requires a different design and evidence:

> Reducing vehicle weight will cause fuel economy to improve by the fitted ALE
> amount in a target population.

AutoXplainR makes the first statement easier to compute and preserve. It does
not turn it into the second.
