---
title: "Bring your own fitted models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bring your own fitted models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

If you already have fitted models, use `evaluate_models()` to open the same report
used by `autoxplain()`. It evaluates the supplied models without fitting,
selecting or replacing them. The primary model is your choice.

```{r evaluate-existing}
training <- mtcars[1:20, ]
evaluation <- mtcars[21:32, ]
fits <- list(
  linear = lm(mpg ~ wt + hp, data = training),
  tree = rpart::rpart(mpg ~ wt + hp, data = training),
  reference = lm(mpg ~ 1, data = training)
)
result <- evaluate_models(
  fits, evaluation, "mpg", features = c("wt", "hp"),
  primary = "linear", reference = "reference"
)
result$leaderboard
predict(result, evaluation[1:3, ])
```

The default data role is `"evaluation"`; the package cannot verify whether these
rows were excluded from fitting or earlier choices. This ordered `mtcars` split
is a teaching example. Declare `evaluation_role = "test"` only when your analysis
design supports that interpretation.

Choose `features` explicitly. By default, every column except the outcome is
treated as an input, including an ID or context column that your adapter might
ignore. The report's data and feature views use this declared input set.

```{r fitted-report}
path <- tempfile(fileext = ".html")
render_model_report(result, path, n_repeats = 5)
file.exists(path)
unlink(path)
```

The report shows scores, fitted behavior, error diagnostics and the supplied data.
Training data are **unavailable** unless you pass `training_data`; supplied
training data add context but do not prove how a model was fitted. No search
history or training time is invented. A reference is optional and can be any
explicitly supplied model. Without one, baseline improvement and paired
reference intervals are unavailable.

The default report embeds aggregate summaries. Set `report_data = "rows"` to add
individual-record inspection and links from errors to source rows. With no
training table supplied, **Explore data** shows only evaluation distributions;
it does not invent a training sample. Save the R
result with `saveRDS()` to retain models, data and custom prediction functions;
external dependencies used by those functions must still be available when you
reload it.

## Custom prediction functions

For an unsupported model class, supply a named `predict_functions` list. Each
function takes `newdata`, or `model, newdata`, and returns one prediction per row.
Regression uses a numeric vector; binary classification uses probabilities of
`positive`; multiclass uses a probability matrix with named class columns.
The outcome factor declares the complete class set. Your adapter owns any
preprocessing needed by the fit; `evaluate_models()` does not learn a recipe.

```{r custom-prediction, eval=FALSE}
result <- evaluate_models(
  list(my_fit = fitted_object), evaluation_data, "outcome",
  features = c("age", "measurement"),
  predict_functions = list(my_fit = function(model, newdata) {
    predict(model, newdata, type = "response")
  }),
  task = "binary", positive = "yes"
)
```

## Work with one explanation or audit

Use `explain_model()` and `audit_explanations()` when you want a focused audit or
an individual estimator rather than the full model report. This interface also
supports hard-label classification for metrics such as accuracy.

```{r existing}
train <- mtcars[1:24, ]
test <- mtcars[25:32, ]
fit <- lm(mpg ~ wt + hp, data = train)
existing <- explain_model(
  fit, test[c("wt", "hp", "mpg")], y = "mpg", label = "existing linear model",
  predict_function = function(model, newdata) as.numeric(predict(model, newdata)),
  metadata = list(evaluation_role = "test")
)
audit <- audit_explanations(existing, n_repeats = 20)
audit$findings
```

A regression prediction function returns one numeric value per row. A binary
function returns the probability of the declared positive class; use `positive`
to name the requested event and `probability_class` when a custom vector describes
the other event. Hard-label adapters can support accuracy, but cannot supply
probability losses or probability effects. Multiclass functions return a probability matrix with
named class columns. Preprocessing must match what the fitted model expects;
`explain_model()` does not learn a new preprocessing recipe. The metadata records
the evaluation role you assert, not independent verification of that role.

For a binomial GLM fitted to numeric 0/1 outcomes, the native response probability
means outcome 1, regardless of evaluation factor order. A factor GLM fitted with
`model = FALSE` no longer retains its original response levels: set
`probability_class` to the event its response probability represents. This follows
[R's binomial response conventions](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/family.html).
Recreate a custom explainer and recompute its evidence after changing external
state used by its prediction function. A content check on observed predictions
cannot establish equivalence on every possible perturbed input.

```{r external-report}
path <- tempfile(fileext = ".html")
render_explanation_report(audit, path)
file.exists(path)
unlink(path)
```

## Narrative from retained evidence

```{r memo}
result <- autoxplain(mtcars, "mpg", seed = 2026)
memo <- generate_natural_language_report(result)
cat(memo)
```

The local memo combines performance, retained importance, audit findings, fitted
effects and failed checks without refitting or rerunning explanations. An explicit
`audit` replaces only that component; `importance_data` and `pdp_data` override
importance and effects respectively. Evaluation evidence is preserved.
Identified audits, effects and importance tables must match the fitted model and
ordered evaluation data. Bare importance tables are labeled user-supplied with
unverified identity; set their `metric` attribute explicitly. A fitting-only
result reports that feature evidence was not supplied.

```{r hosted, eval=FALSE}
# Configure the credentials listed by narrative_providers() first.
memo <- generate_natural_language_report(result, provider = "gemini")
render_model_report(result, "report.html", narrative = memo)
```

Hosted prose requires an explicit provider. The package checks response format
and length and appends interpretation limits; it does not verify numerical
claims or guarantee that the model follows its prompt. Review claims against the
computed output. The default local memo does not send a request, even when API
keys exist. Prompt construction omits raw rows and case predictions, but feature
names, aggregates and diagnostic text may still be sensitive.

Inspect `narrative_providers()` and `?generate_natural_language_report` for adapter
configuration. Provider availability and terms can change independently of the
package.

The memo reads diagnostic status from the supplied result without running new
checks. Rendering may compute optional comparisons in a local copy of that
result; it returns their compact status in `attr(report_path, "diagnostic_status")`
and leaves the original result unchanged. A later memo from the original result
therefore still describes those optional checks as not run unless their status
records were retained.

## Migrating old report calls

Saved 0.5.0 regression, binary and multiclass results were replayed with explicit
explanation recomputation. Their older retained audits are rejected by the newer
identity check. For an unchanged saved fit:

```{r saved-result-migration, eval=FALSE}
older <- readRDS("analysis-0.5.0.rds")
render_model_report(older, "updated-report.html", top_features = 2, n_repeats = 3)
```

The small budget illustrates the migration; choose a suitable explanation budget
for the final report. Rendering validates the available stored predictions and
metrics, then recomputes explanation evidence without refitting. It cannot
recover missing raw context or reconstruct historical external state used by a
custom function on unobserved inputs. Changed models need a new evaluation via
`evaluate_models()` rather than combining their predictions with saved scores.

`generate_dashboard()` and `create_simple_dashboard()` were deprecated in 0.4.0
and remain compatibility entry points in 0.6.0. Replace them with
`render_model_report(result, output_file)`. For a memo, pass
`narrative = generate_natural_language_report(result)`. Each compatibility call
emits one migration warning. Use `evaluate_models()` followed by
`render_model_report()` for existing fitted models, or `render_explanation_report()`
for a standalone explanation audit.
