---
title: "Compare and tune models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare and tune models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

The default, `model_set = "tuned", portfolio = "core"`, schedules 15 settings
across linear, tree and neural families using five training-only folds. It
retains a representative from each successful family and a baseline for the
report. The selected model is evaluated on rows kept outside that search.

`model_set = "quick"` fits a pre-specified linear, logistic or multinomial model
and a baseline. `"comparison"` adds two pre-specified trees but keeps the same
primary model. Its evaluation ranks are descriptive. Use `"tuned"` when the
analysis calls for model selection: candidates are compared using training-only
cross-validation before the selected model is scored on the outer evaluation set.

## Start with regularized models, forests and XGBoost

The unreleased 0.8.0 candidate adds `portfolio = "tabular"` for this combination.
It requires R 4.3 or newer for the supported XGBoost backend; the core package
supports R 4.1 or newer. Install the three optional engines once, then read your
labeled CSV and name its outcome column:

```{r tabular-start, eval=FALSE}
install_model_engines("tabular")
training <- read.csv("train.csv", check.names = FALSE)
result <- autoxplain(training, "outcome", portfolio = "tabular", report = "report.html")
```

Replace `"outcome"` with the target column's name. Before fitting, identify its
task: numeric outcomes with exactly two values select binary classification;
other numeric outcomes select regression. Numeric category codes with three or
more values, such as `0, 1, 2`, need `task = "multiclass"` or conversion to a
factor. Use `task = "regression"` for a numeric two-value regression outcome.
For a binary classification example, set
`training$outcome <- factor(training$outcome, levels = c("no", "yes"))` before
fitting to make `"yes"` the probability target.

`test_data` must contain observed outcomes. Keep a competition's unlabeled test
CSV for prediction after fitting:

```{r tabular-predict, eval=FALSE}
competition_test <- read.csv("test.csv", check.names = FALSE)
predictions <- predict(result, competition_test)
```

The output is a numeric vector for regression, a vector of positive-class
probabilities for binary classification, or a matrix with one probability column
per multiclass outcome. `predict(result, competition_test, type = "class")`
returns class labels for classification.

This starts with 18 candidate settings. On at least 200 outer-training rows,
it screens them on a shared sample and takes the best successful setting from
each family into five-fold cross-validation. It selects the best CV finalist,
refits the retained settings on all training rows and evaluates them on a
separate holdout. The report includes those models and an intercept-only baseline.
Regression, binary classification and multiclass classification use the same call.

Small datasets skip screening and cross-validate each scheduled preset. The
[0.8.0 validation record](https://github.com/Matt17BR/autoXplainR/blob/main/validation/release-0.8.0.md)
reports completed Covertype and Bank comparisons, including forest quality
tradeoffs, measured costs and remaining acceptance work. These results do not
establish competition-level performance. Inspect the fitted results before
deciding what additional modeling work the data need.

## Choose a portfolio

```{r catalog}
as.data.frame(learner_catalog())[, c("family", "backend", "supported_tasks", "portfolios")]
```

| Portfolio | Included families | Dependencies |
|---|---|---|
| `core` | Linear, tree, neural | Ordinary package dependencies |
| `tabular` | Regularized, forest, boosting | `glmnet`, `ranger`, `xgboost` |
| `recommended` | Linear, regularized, additive, tree, forest, boosting | Optional engines installed explicitly |
| `extended` | All ten, including kernel, nearest-neighbor and MARS | Additional optional engines |

Additive and MARS learners support regression and binary tasks. The requested
portfolio does not change silently with installed packages; a missing engine
produces an installation instruction.

```{r tune}
tuned <- autoxplain(iris, "Species", model_set = "tuned", portfolio = "core",
                    max_models = 6, nfolds = 3, seed = 2026)
tuning <- tuning_results(tuned)
tuning$candidates[, c("model", "hyperparameters", "cv_score", "cv_se", "selected")]
```

Every fold learns its preprocessing from that fold's training rows. The core
portfolio's default one-standard-error rule favors the first eligible family in the documented
priority, then its smallest recorded within-family capacity proxy. Some tuning
dimensions are not ordered by that proxy; for example, a neural weight-count
proxy does not order weight decay. Cross-family priority
is a package policy, not a statistical ordering of model families. Fold-score
standard errors are a selection heuristic, not independent-test confidence intervals.
Use `tuning_rule = "best"` for the best resampled score instead. This is already
the default for `portfolio = "tabular"` when `learners` and `tuning_rule` are not
supplied. An explicit `learners` list retains the usual one-standard-error
default; choose `tuning_rule = "best"` when that is the intended policy.

```{r search-evidence}
evidence <- tuning_evidence(tuned)
evidence$selection[c("best_configuration", "selected_configuration", "threshold")]
```

The report's **Model selection** view connects those decisions to the parameter
values, effective fold settings and retained fit. The default grid is a small,
versioned set of contrasting controls. It does not follow a theorem that makes
those numbers optimal for your data. A winning setting on the edge of a searched
range is a reason to investigate that range, not evidence that it should always
be extended. The range includes failed attempts: being inside it does not imply
that neighboring settings produced usable fits. Inspect the successful count
and failure reasons. Candidate comparisons stay inside training data.

Explicit optimizer nonconvergence excludes a configuration by default. Fold and
refit records retain the status and reason. `optimization_policy = "warn"` in
`tuning_control()` deliberately keeps such fits with a warning; an unreported
optimizer status remains unknown. Successful family representatives use their
best valid within-family CV score, while the primary follows the configured
global selection rule. These are different choices.

```{r optional-install, eval=FALSE}
install_model_engines("recommended")
result <- autoxplain(my_data, "outcome", model_set = "tuned", portfolio = "recommended")
```

`tuning_control()` accepts explicit parameter grids, fold IDs, selection metrics
and failure policies. Inspect its help and `tuning_results()` before changing
these defaults. Chronological tuning is not implemented; temporal model selection
requires an explicit rolling-origin design outside this workflow.

## How the tabular search spends its budget

The first stage compares varied settings on one common training/validation
sample of at most 20,000 outer-training rows, with up to 128 forest trees and
600 boosting rounds. The best successful setting in each family receives
complete cross-validation. Large adaptive searches, with at least one million
training rows times predictors, use 256 trees per forest validation fit. At
four million rows times predictors, they use 128. The final all-training forest
has 256 trees in that largest tier and 500 below it. Smaller searches retain
500 trees for both CV and the final fit. Every requested fold and all of its
training rows are retained; the smaller validation forests' scores approximate
the final forest's performance. These tree counts are computation defaults,
not convergence tests. Boosting allows up to 2,000 rounds
and stops after 30 rounds without improvement on its inner score. Screening
scores are shown separately because they describe a different workload.

`search = "auto"` also chooses this adaptive path for an explicit combination
of `regularized`, `forest` and `boosting` that includes a forest or booster,
has at least 200 outer-training rows and has a budget of at least two settings
per family. Custom grids or exact family budgets use `search = "grid"`.
Choose that mode explicitly to give every scheduled configuration full CV.

Boosting chooses its round count on an approximately 80/20 split made inside
each fitting fold, before preprocessing. The inner training rows alone learn
that split's recipe. After stopping, the model is refitted on the whole fitting
fold for scoring. The final model uses all outer-training rows and the rounded-up
median of the successful folds' round counts. If an inner split is skipped,
the fit keeps the requested round budget and records the reason. The final
evaluation rows never select rounds.

Screening uses training outcomes to choose which settings reach full CV.
Those CV scores are therefore **selection evidence conditional on screening**,
not an unbiased accuracy estimate for the whole search procedure. Keep a final
evaluation set separate before choosing the search. Repeated decisions based on
its scores turn it into development data.

Supply `test_data` when you have a separate labeled evaluation set; otherwise
the default holdout comes from the labeled input table.

```{r demanding-search, eval=FALSE}
result <- autoxplain(
  training, "outcome", test_data = final_test, evaluation_role = "test",
  portfolio = "tabular", explain = FALSE, seed = 2026,
  tuning_control = tuning_control(threads = 4, finalists_per_family = 2)
)
search <- tuning_results(result)
search$screening$promotion
search$candidates
result$leaderboard
```

This example takes up to two successful settings per family into full CV,
instead of the default one. It explicitly requests four CPU threads in each
native forest or boosting fit. Automatic adaptive searches with at least one
million training rows times predictors use up to four available cores; smaller
and exact searches use one. The automatic allocation respects process and job
limits through `parallelly::availableCores()`. An explicit count overrides it.
Candidate fits still run sequentially, and replay records the resolved count.
Increasing finalists spends more time checking the screening decision.
Increasing `max_models` instead proposes more settings to screen.

Tuned forest and boosting workflows show their current stage, setting and fold
automatically from 200 input rows. During longer importance calculations, updates
count successful shuffles at most once every 30 seconds. An active native fit or
prediction finishes before an update can appear. Set `verbosity = "quiet"` to
silence progress or `verbosity = "info"` to enable it on smaller inputs.

Use `tuning_control(time_limit = 600)` to stop scheduling additional search work
after ten minutes. The initial screening round and the first usable candidate
are allowed to complete, and an active fit is not interrupted. Final refits,
explanations and report generation are outside this budget. The record
distinguishes a failed fit, a screened-out setting and a setting skipped because
the budget was used. `time_limit` is a scheduling control, not a hard timeout.

The report's **Model selection** tab shows promotion decisions, actual settings,
fold scores, round choices and failure reasons. In R, inspect `search$screening`,
`search$fold_scores` and `search$refit`. Settings cover plausible alternatives;
they are not an exhaustive search or a guarantee that the best model was found.
The [current comparison protocol](https://github.com/Matt17BR/autoXplainR/blob/main/validation/competitive-tabular/README.md)
tests the development workflow against native forests and XGBoost on separate
development and acceptance data.

## Choose the score that matches the problem

Set the metric before fitting. It governs candidate selection and, when enabled,
the stopping decision for boosting. It also becomes the main held-out report score.

| Task | `tuning_control(metric = ...)` | Interpretation |
|---|---|---|
| Regression | `"rmse"` (default), `"mae"` | Prediction error in the outcome's units; lower is better |
| Nonnegative regression | `"rmsle"` | Root mean squared error after `log1p()`; lower is better |
| Binary or multiclass | `"log_loss"` (default), `"brier"` | Probability error; lower is better |
| Binary | `"auc"` | Positive-negative ranking, with half credit for ties; higher is better |

```{r metric-choice, eval=FALSE}
result <- autoxplain(training, "outcome", portfolio = "tabular",
                     tuning_control = tuning_control(metric = "auc"))
```

Binary probabilities refer to the second outcome factor level. Set the levels
explicitly, for example `factor(outcome, levels = c("no", "yes"))`.
AUC is recorded as `roc_auc`. Its CV score averages within-fold AUC, weighted
by each fold's evaluated row count. It does not rank predictions from different
fold models together. AUC has no additive per-row loss, so its out-of-fold `case_loss`
is `NA`; probabilities and class labels remain available. The fold SE is only
a selection heuristic, not a confidence interval for AUC.

RMSLE requires nonnegative outcomes and predictions. It does not transform the
training target or clip negative predictions. A candidate with a negative
validation prediction cannot receive a valid RMSLE score. If a shuffle produces
a negative prediction, that feature's importance is unavailable and the report
explains why; the other features can still be inspected.

Failed configurations remain in the search record. A completed model can still
be unsuitable: compare its held-out score with the baseline and inspect its
mistakes. AUC assesses ranking; log loss and calibration assess probability
quality. A good result on one does not imply a good result on the others.

## Budget explanations and the report separately

Fitting, explaining and exporting are different costs. `explain = FALSE` lets you
inspect the search first. Request a small report explicitly, then increase its
explanation budget when the fitted models deserve closer investigation:

```{r bounded-report, eval=FALSE}
render_model_report(result, "first-look.html", top_features = 3, n_repeats = 5,
                    explanation_rows = 2000, report_data = "summary",
                    uncertainty = FALSE)
```

Even a small feature audit first screens inputs; expensive prediction functions
and hundreds of columns can take time. Five permutations are a first look, not
precise importance estimates. The report records the budget and warns about
unstable Monte Carlo estimates.

This first report also omits paired score intervals. By default, those intervals
use 1,000 bootstrap resamples of the complete evaluation set, independently of
the explanation and export limits. `uncertainty = FALSE` skips that calculation;
model scores still use the complete evaluation set. Omit this argument to
include the intervals when the validation design supports them.

Rendering returns the HTML path and leaves `result` unchanged. If the result was
fitted with `explain = FALSE`, repeat the explanation settings in later renders:
the first report does not save its evidence or budget back into `result`.
Omitting those settings computes default explanations unless the result already
contains retained explanations.

## Larger data

The rows used to fit and score models are distinct from the rows used to explain
them or included in HTML. `autoxplain()` fits on the complete processed training
partition and scores the complete processed evaluation partition. Rows removed
by an explicit preprocessing rule remain recorded as omissions.

The report's default importance and effect calculations use at most 5,000
evaluation rows. Pairwise plots and associations use at most 10,000 rows per
partition. Unfiltered distributions and missing counts use all available rows.
Filtered charts and their counts describe only matching exported records;
sidebar counts keep the full population.
Official model scores always use the complete evaluation partition and are
unchanged by filters. Each sampled view gives its own denominator. A rare
class or small cluster can be absent from a uniform sample even when its full
count is visible elsewhere. Increasing permutation repeats does not resolve
that absence, and shuffle intervals do not measure row-sampling uncertainty.

```{r larger-data-controls, eval=FALSE}
result <- autoxplain(
  training, "outcome", test_data = final_test,
  learners = c("regularized", "boosting"), max_models = 12, nfolds = 3,
  tuning_control = tuning_control(retain_oof = FALSE),
  explanation_rows = 5000, report = "model-report.html"
)

# Include a bounded sample for record inspection and browser filters.
render_model_report(
  result, "records.html",
  report_data = report_data_control(mode = "rows", max_rows = 20000, max_pair_rows = 10000)
)
```

Choose the search and validation design before looking at the final test scores.
The example budget is a starting point, not a claim that twelve settings are
enough for a difficult problem. `retain_oof = FALSE` saves memory by omitting
case-level CV predictions; fold scores, candidate settings and selection remain
available. Use the default `TRUE` when you need to inspect individual CV errors.
This control does not reduce fitting rows or final evaluation rows.

The 0.7.0 million-row regression control used a smaller fixed search: one
regularized and one boosting configuration, two CV folds and `retain_oof = FALSE`.
One call fitted all million training rows, scored all 20,000 independent
evaluation rows and wrote a 3,891,782-byte summary report in 81.055 seconds.
Default explanations used 5,000 rows; all 1,000 paired bootstrap draws used the
full evaluation set. The shallow selected model had RMSE 1.6835. A separate
three-configuration search selected 600 boosting rounds at depth 6 and reached
RMSE 0.7491 in 109.012 seconds with explanations disabled. These single-host
results measure different workloads; see the
[measurements](https://github.com/Matt17BR/autoXplainR/blob/main/validation/scalability/findings.md).

Large result checks write complete model and data state to a temporary file,
then remove it. The temporary file system needs space for that serialization.
If you need disk-backed temporary storage, set `TMPDIR` to an existing, writable
directory on disk before starting R. Changing it after R starts does not move
the session's temporary directory.

Neural configurations now allow up to 2,000 optimizer iterations, and stop
earlier when they converge. The former 500-iteration ceiling excluded wider
networks on the nonlinear stress problem even when more iterations produced
useful fits. `maxit` is an explicit neural grid setting, so you can retain the
former budget or choose another allowance. Inspect convergence and failed
configurations alongside scores. A larger allowance can improve model selection
but also increases fitting time; it does not guarantee convergence.

Use `explanation_rows = NULL` to remove the report explanation cap and
`report_data_control(max_pair_rows = NULL)` for all-row pair summaries.
The lower-level `calculate_permutation_importance()`, `audit_explanations()` and
`explain_effect()` have no shared row cap by default and accept `max_rows` explicitly.
PDP has its separate `sample_size` control: its curve and support distribution
can use different row counts, both recorded in the result.

Large reports store compressed column values and unpack them offline as needed.
Compression does not remove exported values. The Data tab's linked scatter shows
at most 1,500 points; filters and record lookup use all exported rows. Exporting every
row can still create a large file and considerable browser memory use; leave
`report_data = "summary"` unless individual records serve a concrete purpose.

For large categorical expansions, automatic boosting encoding uses native
categorical splits instead of a numeric contrast matrix. The threshold is an
estimated matrix with more than 50 million cells and more than twice the input
width. The representation is fixed for a search from outer-training inputs;
category mappings and preprocessing stay inside each fold. The two
representations are different models, not interchangeable numerical shortcuts.
Set `encoding = "matrix"` or `"native"` in an explicit boosting grid to control
this choice. The search record and report show what was used and why.

Use `report_data_control()` to restrict exported columns or sample records if
you need row filtering. Full records for hundreds of columns make a large HTML
file. Aggregate mode keeps full univariate summaries and bounded pair summaries
but does not allow row filtering.
`max_models` bounds the candidate pool, not elapsed time. Adaptive screening and
early-stopping calibration can fit a candidate more than once.
`max_runtime_secs` is an H2O setting and does not impose a local-engine timeout.

## Compare fitted behavior

```{r behavior}
comparison <- autoxplain(iris, "Sepal.Length", model_set = "comparison", seed = 2026)
compare_model_behavior(comparison)
head(prediction_ambiguity(comparison)$rows)
```

Family descriptions state what an algorithm can represent. The scores and
prediction differences describe what these particular fitted models did on
common evaluation rows. Disagreement does not supply prediction intervals or
identify the correct prediction.

```{r effects, fig.width=7, fig.height=4}
effects <- compare_model_effects(comparison, "Petal.Length", method = "ale", n_points = 8)
plot(effects)
```

Effects are evaluated on a common grid. Their differences cover the supplied
fits, not every competitive model or uncertainty from refitting. For multiclass
outcomes, request a named class with `class = "virginica"`.

`model_tradeoffs()` offers optional performance/resource comparisons. Its default
cost uses a recorded repeated prediction benchmark when available, then retained
fit time, evaluation-batch prediction time, and R object size. It requires at
least two finite score/cost pairs. Fit time excludes the search; R size includes
retained diagnostics and excludes native allocations. Use these measurements
for a concrete resource constraint; a favorable
Pareto position is not a model-selection rule.

For repeated prediction measurements on the same rows:

```{r prediction-cost, eval=FALSE}
bench <- benchmark_predictions(comparison)
render_model_report(comparison, "report.html", benchmark = bench)
```

The benchmark records a common batch, warmup, repeated blocks, warnings and
timing limits. A per-row batch cost is not single-request latency;
quartiles describe the repeated measurements, not a confidence interval. The
benchmark excludes fitting and the guided workflow's raw-data preprocessing.
Transformations performed inside a custom prediction function are timed with
that function.

## Optional H2O search

H2O is a separate engine requiring Java, the `h2o` package and a running process.
It does not support this package's grouped or temporal validation plans. With
H2O internal cross-validation, learned external imputation, missingness-based
column removal and automatic ordinal coercion are rejected because they would
be learned before H2O creates its folds. Use the default keep-missing path, the
base engine for fold-local preprocessing, or explicitly designed validation
with `nfolds = 0`. H2O's own transformations remain engine responsibilities.

```{r h2o, eval=FALSE}
holdout <- seq(5L, nrow(iris), by = 5L)
h2o_result <- autoxplain(
  iris[-holdout, ], "Species", test_data = iris[holdout, ],
  engine = "h2o", evaluation_role = "test", max_models = 10,
  max_runtime_secs = 0, exclude_algos = "DeepLearning", seed = 2026
)
render_model_report(h2o_result, "h2o-report.html")
```

This keeps supplied evaluation rows out of H2O selection. A fixed model budget,
no wall-clock limit and exclusion of Deep Learning reduce known sources of H2O
variation; consult the recorded provenance for remaining limits. See the
[statistical methods](statistical-methods.html) for evaluation assumptions.
