---
title: "Compare and tune models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare and tune models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

The default, `model_set = "tuned", portfolio = "core"`, schedules 15 settings
across linear, tree and neural families using five training-only folds. It
retains a representative from each successful family and a baseline for the
report. The selected model is evaluated on rows kept outside that search.

`model_set = "quick"` fits a pre-specified linear, logistic or multinomial model
and a baseline. `"comparison"` adds two pre-specified trees but keeps the same
primary model. Its evaluation ranks are descriptive. Use `"tuned"` when the
analysis calls for model selection: candidates are compared using training-only
cross-validation before the selected model is scored on the outer evaluation set.

## Choose a portfolio

```{r catalog}
as.data.frame(learner_catalog())[, c("family", "backend", "supported_tasks", "portfolios")]
```

| Portfolio | Included families | Dependencies |
|---|---|---|
| `core` | Linear, tree, neural | Ordinary package dependencies |
| `recommended` | Linear, regularized, additive, tree, forest, boosting | Optional engines installed explicitly |
| `extended` | All ten, including kernel, nearest-neighbor and MARS | Additional optional engines |

Additive and MARS learners support regression and binary tasks. The requested
portfolio does not change silently with installed packages; a missing engine
produces an installation instruction.

```{r tune}
tuned <- autoxplain(iris, "Species", model_set = "tuned", portfolio = "core",
                    max_models = 6, nfolds = 3, seed = 2026)
tuning <- tuning_results(tuned)
tuning$candidates[, c("model", "hyperparameters", "cv_score", "cv_se", "selected")]
```

Every fold learns its preprocessing from that fold's training rows. The default
one-standard-error rule favors the first eligible family in the documented
priority, then its smallest recorded within-family capacity proxy. Some tuning
dimensions are not ordered by that proxy; for example, a neural weight-count
proxy does not order weight decay. Cross-family priority
is a package policy, not a statistical ordering of model families. Fold-score
standard errors are a selection heuristic, not independent-test confidence intervals.
Use `tuning_rule = "best"` for the lowest resampled error instead.

```{r search-evidence}
evidence <- tuning_evidence(tuned)
evidence$selection[c("best_configuration", "selected_configuration", "threshold")]
```

The report's **Model selection** view connects those decisions to the parameter
values, effective fold settings and retained fit. The default grid is a small,
versioned set of contrasting controls. It does not follow a theorem that makes
those numbers optimal for your data. A winning setting on the edge of a searched
range is a reason to investigate that range, not evidence that it should always
be extended. The range includes failed attempts: being inside it does not imply
that neighboring settings produced usable fits. Inspect the successful count
and failure reasons. Candidate comparisons stay inside training data.

Explicit optimizer nonconvergence excludes a configuration by default. Fold and
refit records retain the status and reason. `optimization_policy = "warn"` in
`tuning_control()` deliberately keeps such fits with a warning; an unreported
optimizer status remains unknown. Successful family representatives use their
lowest valid within-family CV loss, while the primary follows the configured
global selection rule. These are different choices.

```{r optional-install, eval=FALSE}
install_model_engines("recommended")
result <- autoxplain(my_data, "outcome", model_set = "tuned", portfolio = "recommended")
```

`tuning_control()` accepts explicit parameter grids, fold IDs, selection rules
and failure policies. Inspect its help and `tuning_results()` before changing
these defaults. Chronological tuning is not implemented; temporal model selection
requires an explicit rolling-origin design outside this workflow.

## Compare fitted behavior

```{r behavior}
comparison <- autoxplain(iris, "Sepal.Length", model_set = "comparison", seed = 2026)
compare_model_behavior(comparison)
head(prediction_ambiguity(comparison)$rows)
```

Family descriptions state what an algorithm can represent. The scores and
prediction differences describe what these particular fitted models did on
common evaluation rows. Disagreement does not supply prediction intervals or
identify the correct prediction.

```{r effects, fig.width=7, fig.height=4}
effects <- compare_model_effects(comparison, "Petal.Length", method = "ale", n_points = 8)
plot(effects)
```

Effects are evaluated on a common grid. Their differences cover the supplied
fits, not every competitive model or uncertainty from refitting. For multiclass
outcomes, request a named class with `class = "virginica"`.

`model_tradeoffs()` offers optional performance/resource comparisons. Its default
size axis measures approximate R object storage, not structural complexity.
Use it when storage or runtime is relevant to a concrete constraint; a favorable
Pareto position is not a model-selection rule.

For repeated prediction measurements on the same rows:

```{r prediction-cost, eval=FALSE}
bench <- benchmark_predictions(comparison)
render_model_report(comparison, "report.html", benchmark = bench)
```

The benchmark records a common batch, warmup, repeated blocks, warnings and
timing limits. A per-row batch cost is not single-request latency;
quartiles describe the repeated measurements, not a confidence interval. The
benchmark excludes fitting and the guided workflow's raw-data preprocessing.
Transformations performed inside a custom prediction function are timed with
that function.

## Optional H2O search

H2O is a separate engine requiring Java, the `h2o` package and a running process.
It does not support this package's grouped or temporal validation plans. With
H2O internal cross-validation, learned external imputation, missingness-based
column removal and automatic ordinal coercion are rejected because they would
be learned before H2O creates its folds. Use the default keep-missing path, the
base engine for fold-local preprocessing, or explicitly designed validation
with `nfolds = 0`. H2O's own transformations remain engine responsibilities.

```{r h2o, eval=FALSE}
holdout <- seq(5L, nrow(iris), by = 5L)
h2o_result <- autoxplain(
  iris[-holdout, ], "Species", test_data = iris[holdout, ],
  engine = "h2o", evaluation_role = "test", max_models = 10,
  max_runtime_secs = 0, exclude_algos = "DeepLearning", seed = 2026
)
render_model_report(h2o_result, "h2o-report.html")
```

This keeps supplied evaluation rows out of H2O selection. A fixed model budget,
no wall-clock limit and exclusion of Deep Learning reduce known sources of H2O
variation; consult the recorded provenance for remaining limits. See the
[statistical methods](statistical-methods.html) for evaluation assumptions.
