---
title: "Compare and tune models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare and tune models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

`model_set = "quick"` fits a pre-specified linear, logistic or multinomial model
and a baseline. `"comparison"` adds two pre-specified trees but keeps the same
primary model. Its evaluation ranks are descriptive. Use `"tuned"` when the
analysis calls for model selection: candidates are compared using training-only
cross-validation before the selected model is scored on the outer evaluation set.

## Choose a portfolio

```{r catalog}
as.data.frame(learner_catalog())[, c("family", "backend", "supported_tasks", "portfolios")]
```

| Portfolio | Included families | Dependencies |
|---|---|---|
| `core` | Linear, tree, neural | Ordinary package dependencies |
| `recommended` | Linear, regularized, additive, tree, forest, boosting | Optional engines installed explicitly |
| `extended` | All ten, including kernel, nearest-neighbor and MARS | Additional optional engines |

Additive and MARS learners support regression and binary tasks. The requested
portfolio does not change silently with installed packages; a missing engine
produces an installation instruction.

```{r tune}
tuned <- autoxplain(iris, "Species", model_set = "tuned", portfolio = "core",
                    max_models = 6, nfolds = 3, seed = 2026)
tuning <- tuning_results(tuned)
tuning$candidates[, c("model", "hyperparameters", "cv_score", "cv_se", "selected")]
```

Every fold learns its preprocessing from that fold's training rows. The default
one-standard-error rule favors the first eligible family in the documented
priority, then its least-flexible eligible configuration. Cross-family priority
is a package policy, not a statistical ordering of model families. Fold-score
standard errors are a selection heuristic, not independent-test confidence intervals.
Use `tuning_rule = "best"` for the lowest resampled error instead.

```{r optional-install, eval=FALSE}
install_model_engines("recommended")
result <- autoxplain(my_data, "outcome", model_set = "tuned", portfolio = "recommended")
```

`tuning_control()` accepts explicit parameter grids, fold IDs, selection rules
and failure policies. Inspect its help and `tuning_results()` before changing
these defaults. Chronological tuning is not implemented; temporal model selection
requires an explicit rolling-origin design outside this workflow.

## Compare fitted behavior

```{r behavior}
comparison <- autoxplain(iris, "Sepal.Length", model_set = "comparison", seed = 2026)
compare_model_behavior(comparison)
head(prediction_ambiguity(comparison)$rows)
```

Family descriptions state what an algorithm can represent. The scores and
prediction differences describe what these particular fitted models did on
common evaluation rows. Disagreement does not supply prediction intervals or
identify the correct prediction.

```{r effects, fig.width=7, fig.height=4}
effects <- compare_model_effects(comparison, "Petal.Length", method = "ale", n_points = 8)
plot(effects)
```

Effects are evaluated on a common grid. Their differences cover the supplied
fits, not every competitive model or uncertainty from refitting. For multiclass
outcomes, request a named class with `class = "virginica"`.

`model_tradeoffs()` offers optional performance/resource comparisons. Its default
size axis measures approximate R object storage, not structural complexity.
Use it when storage or runtime is relevant to a concrete constraint; a favorable
Pareto position is not a model-selection rule.

## Optional H2O search

H2O is a separate engine requiring Java, the `h2o` package and a running process.
It does not support this package's grouped or temporal validation plans. With
H2O internal cross-validation, learned external imputation, missingness-based
column removal and automatic ordinal coercion are rejected because they would
be learned before H2O creates its folds. Use the default keep-missing path, the
base engine for fold-local preprocessing, or explicitly designed validation
with `nfolds = 0`. H2O's own transformations remain engine responsibilities.

```{r h2o, eval=FALSE}
holdout <- seq(5L, nrow(iris), by = 5L)
h2o_result <- autoxplain(
  iris[-holdout, ], "Species", test_data = iris[holdout, ],
  engine = "h2o", evaluation_role = "test", max_models = 10,
  max_runtime_secs = 0, exclude_algos = "DeepLearning", seed = 2026
)
render_model_report(h2o_result, "h2o-report.html")
```

This keeps supplied evaluation rows out of H2O selection. A fixed model budget,
no wall-clock limit and exclusion of Deep Learning reduce known sources of H2O
variation; consult the recorded provenance for remaining limits. See the
[statistical methods](statistical-methods.html) for evaluation assumptions.
