---
title: "Choose validation and inspect diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choose validation and inspect diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(AutoXplainR)
```

Choose the sampling unit before interpreting scores. A random row holdout
assumes rows can be sampled independently. Repeated people, households or sites
usually need whole-group partitions; observations over time need a chronological
boundary and features available at prediction time.

```{r designs, eval=FALSE}
result <- autoxplain(measurements, "outcome",
                     validation = validation_split(group = "site_id"))
result <- autoxplain(observations, "outcome",
                     validation = validation_split(time = "recorded_at", gap = 2))
```

Split columns are excluded from predictors. Fractions refer to whole groups or
distinct times, so row fractions can differ. A temporal gap of two excludes two
distinct time values, not necessarily two days. Tied times stay together.
Grouped tuning keeps units intact and fails if classification folds cannot
contain every outcome class. Chronological tuning is unsupported.

For an external evaluation set, supply `test_data`. The default role is
`"evaluation"`; use `evaluation_role = "test"` only when the study design supports
that role. Duplicate-valued rows across training and evaluation produce a possible
leakage warning. The role label and duplicate check cannot establish independence.

## Separate different uncertainties

```{r uncertainty}
result <- autoxplain(mtcars, "mpg", explain = FALSE)
intervals <- performance_uncertainty(result, n_boot = 100, seed = 2026)
intervals$estimates
```

This small teaching example uses 100 bootstrap draws; use at least 1000 for an
analysis. The paired bootstrap measures evaluation-sample variation conditional
on the fitted primary model and baseline. Grouped designs resample whole groups.
It does not include uncertainty from fitting or model selection, and temporal
designs reject the IID bootstrap. See [statistical methods](statistical-methods.html).

## Probabilities, thresholds and groups

```{r classification}
cars <- transform(mtcars, am = factor(am, labels = c("automatic", "manual")))
fit <- autoxplain(cars, "am")
calibration_diagnostics(fit)
threshold_diagnostics(fit)
```

Binary probabilities refer to the second training factor level, here `"manual"`.
Calibration compares predicted probabilities with observed frequencies. The
binned calibration gap depends on the sample and chosen grouping. Threshold
sensitivity displays false-positive/false-negative trade-offs; choosing a
threshold after viewing final evaluation labels would reuse those labels for selection.

```{r subgroup}
regression <- autoxplain(cars, "mpg")
subgroup_performance(regression, "am")
```

Subgroup scores describe an explicitly selected column. They are not a fairness
certification, and small groups may supply insufficient evidence. The function
retains group counts and reasons for unavailable metrics.

## Missingness changes

```{r missingness}
set.seed(23)
training <- data.frame(x = rnorm(100), y = rnorm(100))
evaluation <- data.frame(x = rnorm(40), y = rnorm(40))
evaluation$x[1:12] <- NA_real_
missing_fit <- autoxplain(training, "y", test_data = evaluation, explain = FALSE)
missingness_shift(missing_fit)
```

Missingness rates are recorded before imputation. A difference is a descriptive
review signal, not a distribution-shift test or proof that imputation preserved
predictive validity. The report and result retain unavailable checks separately
from completed checks.
